---
title: "Geography 176A"
author: "[Mika Leslie](https://mikaleslie.github.io)"
subtitle: 'Lab 02: COVID-19 Pandemic'
output:
  html_document:
    theme: journal
---

###Question 1

```{r, warning=FALSE, message=FALSE, echo=FALSE}
library(tidyverse)
library(readxl)
library(zoo)
library(knitr)
library(ggplot2)
covid = read_csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv")

state.of.interest = "California"

covid %>%
  filter(state == state.of.interest) %>%
  group_by(county) %>%
  mutate(newCases = cases - lag(cases)) %>%
  ungroup() ->
  newCases2
```

####Table 1

```{r, warning=FALSE, message=FALSE, echo = FALSE}
newCases2 %>%
  filter(date == max(date)) %>%
  arrange(-cases) %>%
  head(5) %>%
  select(county, cases) %>%
  knitr::kable(caption = paste("Most Cumulative Cases", state.of.interest),
               col.names = c("County", "Cumulative Cases"),
               format.args = list(big.mark = ",")) %>%
  kableExtra::kable_styling("bordered", full_width = F, font_size = 14)
```

####Table 2
```{r, echo = FALSE}
newCases2 %>%
  filter(date == max(date)) %>%
  arrange(-newCases) %>%
  head(5) %>%
  select(county, newCases) %>%
  knitr::kable(caption = paste("Most New Cases", state.of.interest),
               col.names = c("County", "New Cases"),
               format.args = list(big.mark = ",")) %>%
  kableExtra::kable_styling("bordered", full_width = F, font_size = 14)
```


```{r, echo = FALSE}
#Part 4, 5
pop <- read_excel("~/github/geog-176A-labs/data/PopulationEstimates.xls", skip = 2) %>%
  select (fips = "FIPStxt", state = "State", Area_Name, pop2019 = "POP_ESTIMATE_2019")
```

Exploring the data
```{r, eval = TRUE}
names(pop)
nrow(pop)
str(pop)
dim(pop)
```


```{r, echo = FALSE}
#part 7
covid_pop = left_join(newCases2, pop, by = "fips")
```

####Table 3
```{r, echo = FALSE}
#part 8
covid_pop %>%
  filter(date == max(date)) %>%
  mutate(cases_pcap = cases/pop2019) %>%
  arrange(-cases_pcap) %>%
  head(5) %>%
  select(county, cases_pcap) %>%
  knitr::kable(caption = paste("Most Cumulative Cases Per Capita", state.of.interest),
               col.names = c("County", "Cumulative Cases Per Capita"),
               format.args = list(big.mark = ",")) %>%
  kableExtra::kable_styling("bordered", full_width = F, font_size = 14)
```

####Table 4
```{r, warning=FALSE, message=FALSE, echo = FALSE}
covid_pop %>%
  filter(date == max(date)) %>%
  mutate(newcases_pcap = newCases/pop2019) %>%
  arrange(-newcases_pcap) %>%
  head(5) %>%
  select(county, newcases_pcap) %>%
  knitr::kable(caption = paste("Most New Cases Per Capita", state.of.interest),
               col.names = c("County", "New Cases Per Capita"),
               format.args = list(big.mark = ",")) %>%
  kableExtra::kable_styling("bordered", full_width = F, font_size = 14)
```


```{r, echo = FALSE}
covid_pop %>%
  filter(date >= max(date)-13) %>%
  group_by(county, pop2019) %>%
  summarize(newCases_p100 =  sum(newCases)) %>%
  mutate(newCases_p100 = newCases_p100 / (pop2019/100000)) %>%
  filter(newCases_p100 < 100) ->
  final

newCases2 %>%
  summarise(tot_cases = sum(cases)) -> x

newCases2 %>%
  summarise(tot_newcases = sum(newCases, na.rm = T)) -> y

paste("The total number of cases is", x,
      "the total number of new cases is", y,
      "and the total number of safe counties is", nrow(final))
```

###Question 2

```{r, warning=FALSE, message=FALSE, echo = FALSE}
covid = read_csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv")

#part 2
covid %>%
  filter(state == "California" | state == "New York" | state == "Florida" | state == "Louisiana") %>%
  group_by(state, date) %>%
  summarise(cases = sum(cases)) %>%
  mutate(newCases = cases - lag(cases),
         roll7 = rollmean(newCases, 7, fill = NA, align="right")) %>%
  ungroup() %>%
  ggplot(aes(x = date))+
  geom_col(aes(y = newCases, color = state)) +
  geom_line(aes(y = roll7), size = 1) +
  facet_wrap(~state)+
  theme_dark() +
  labs(title = paste("New Reported Cases by Day in States of Interest"),
       color = "",
       x = "Date",
       y = "") +
  theme(plot.background = element_rect(fill = "white"),
        panel.background = element_rect(fill = "white"),
        plot.title = element_text(size = 14, face = 'bold'))
```



```{r, warning=FALSE, message=FALSE, echo = FALSE}
pop <- read_excel("~/github/geog-176A-labs/data/PopulationEstimates.xls", skip = 2) %>%
  select (fips = "FIPStxt", state = "Area_Name", pop2019 = "POP_ESTIMATE_2019")

covid_pop = left_join(covid, pop, by = "state")

covid_pop %>%
  filter(state == "California" | state == "New York" | state == "Florida" | state == "Louisiana") %>%
  select(date, state, cases, pop2019) %>%
  group_by(state, date, pop2019) %>%
  summarise(cases = sum(cases)) %>%
  ungroup() %>%
  group_by(state) %>%
  mutate(newCases = cases - lag(cases),
         newCases_pcap = newCases/pop2019,
         roll7 = rollmean(newCases_pcap, 7, fill = NA, align="right")) %>%
  ggplot(aes(x = date))+
  geom_col(aes(y = newCases_pcap, color = state)) +
  geom_line(aes(y = roll7), size = 1) +
  facet_wrap(~state)+
  theme_dark() +
  labs(title = paste("New Reported Cases Per Capita by Day in States of Interest"),
       color = "",
       x = "Date",
       y = "") +
  theme(plot.title = element_text(size = 10, face = 'bold'))
```

By looking at the number of cases per capita, some states appear to do
better than others compared to the raw data. The peak in the CA plot appears
much smaller, while the peak in the LA plot appears much larger because
CA's population is much larger. So while CA has more cases than LA, they have
a smaller ratio to the total population.
